Stack Safety as a Security Property

### 2020-07-09 main.tex
Overall pass, some reworking of policy discussions

The intuition behind \citeauthor{Skorstengaard+19}'s use of the term {\em
  local state encapsulation} is that callers should be protected from
callees: a callee should be not be able to access or overwrite private data
in the caller's stack frame. We formalize this intuition as a security
property with both confidentiality and integrity components.
%
We reflect this intuition in a formal
{\em stack confidentiality} property, which promises that a callee's
observable behavior is not affected by memory outside its stack frame
(e.g., a caller's private data or uninitialized memory).  Conversely, {\em stack
  integrity} promises that, if a function call ever returns, the caller's
stack data at the point of return will be identical to what it was before
the call.

@@

To phrase these two aspects of stack safety in terms of security
properties \leo{in terms of security properties or ``using security
  language''?}  requires an explicit attacker model: an clear
understanding of who is being protected and from what. Most existing
stack protection mechanisms deal with scenarios where a callee gets
influenced by external inputs and ends up executing potentially
malicious code. To avoid the details of exactly how this happens and
keep the presentation clear, we pick an even stronger model: when
stating the protections that a function can rely on when it makes a
call, we assume that the callee can execute arbitrary instructions
until it returns.

@@

This model is flexible enough to capture any kind of mechanism used for dynamic
enforcement of properties of interest, including the following representative
families of policies.

@@

\paragraph*{Null policy}
%
The simplest possible policy is the neutral policy that carries no information
and does nothing. Its set of policy states $\POLS$ is the unit type, and its
policy step function is the constant function that produces the unit, $\lambda
\_ . \top$, thereby allowing all calls the machine step function. Its
initialization function is defined on all states. This corresponds to the
lifting of our machine model to the policy setting. Machines that fulfill a
property by construction will also fulfill that property with the null
policy.

\paragraph*{Software-only policies}
%
Some policies are implemented in software, by modifying code in the compiler
to (e.g.) perform bounds checking \citep{NagarakatteZMZ09} or to insert stack
canaries \citep{Cowan+98}. These approaches are represented as policies that
accept only initial machine states that are well-formed for the policy (due
to being produced by the compiler.) The set of policy states \(\POLS\) is again
the unit type and the step function constant. Property enforcement comes from
the fact that only ill-formed states can induce a trace that violates the
property.

@@

\paragraph*{Capability machines}
%
Other policies are implemented in hardware or built on such hardware mechanisms.
Capability machines like CHERI \rb{Woodruff+14} extend conventional
architectures to support efficient and fine-grained control over memory
accesses, which can be used to implement security policies involving, for
example, compartmentalization or the stack protection policies of
\citet{Skorstengaard+19,Skorstengaard+19b}. Mapping the underlying protection
mechanisms to policy states and step function would be a delicate operation, as
those are tightly coupled to the hardware design. However, separating the
enforcement aspects of a concrete policy (such as StkTokens, where \(\POLS\)
would encompass details like the abstract call stack, the capability sealing
scheme, etc. \rb{is this digging too deep? or on the contrary, say more?}) from
the machine would be relatively straightforward.

@@

In this section we describe our machine model. Our definition of stack
safety is generic and this section discusses an abstract interface to
our machine model and some kind of static or dynamic, software- or
hardware-enforced security monitor. We will refer back to the running
example of the previous section for concreteness, and then,
in~\cref{sec:enforcement}, extend this model with a dynamic policy
enforcement mechanism based on
\citet{DBLP:conf/sp/RoesslerD18}.\apt{needs a relook}

@@

In our model, the machine step function is total -- a program does not halt
on its own, but may instead enter a silent terminal loop. Runs are only cut
short by the policy step function. This means that software-only policies that
can catch errors must be modeled as reaching an error state\apt{??}, not 
halting outright.\apt{Is this our current thinking?}

@@

In this section we formalize our definition of stack safety.We
begin with our model of \leo{compiler-provided?} knowledge about the
program structure, and then dive into the formal definitions of local state
encapsulation and well-bracketed control flow that together constitute
stack safety.

@@

\definition%{Stack Confidentiality}
%Let $\callmap$ be a call map and $\codemap$ a function map.
We will say that a system satisfies {\em stack confidentiality} with respect to
\(\callmap\) in a similar way to stack integrity. For every initial state
\(\mach_0\), we take the induced trace \(\mpstate{\mach_0}{\pinit(\mach_0)}
\hookrightarrow \MPT\). Then for any \(n\)-transition
\(\mathit{Tr}_n ~ \MPT ~ \mpstate{\mach_c}{\pol_c} ~
\mpstate{\mach_e}{\pol_e}\), we take the prefix trace until return:
%
\[\mpstate{\mach_e}{\pol_e} \hookrightarrow \MPT_{call} | \ret{\mach_c}\]

We also take the corresponding trace from any variant \(\mach_e' 
\approx_{\mathit{Cof}(\mach_c,n)} \mach_e\):
%
\[\mach_e' \hookrightarrow \machT_{call} | \ret{\mach_c}\]

@@

@@

\paragraph*{Micropolicies}
%
The enforcement policy of \citeauthor{DBLP:conf/sp/RoesslerD18} relies
on programmable, tag-based reference monitor that runs alongside the
program.  To control this monitor, they use a programming model that
allows fine-grained manipulation of metadata tags to encode so-called
\emph{micropolicies}~\citep{pump_oakland15}.

In such a system, all memory addresses and registers (including the
program counter) are enriched with an abstract metadata tag, which can
point to the representation of an arbitrary data structure. A
micropolicy is thus defined by a collection of rules to be checked as
each instruction executes and together define the transfer function of
the security policy. Each rule inspects the tags associated to the
inputs of an instruction (PC, registers, memory) as well as the
instruction opcode itself, and produces one of two outcomes: either
the instruction is allowed to execute (and generates and assigns tags
for the new PC and the result of the operation), or the machine
fail-stops with a policy violation.

Efficient execution of these micropolicies relies on hardware
implementations, such as the PUMP architecture~\citep{pump_oakland15},
which include a monitor rule cache that maps tag inputs to outputs and
a mechanism to trap misses to software handlers and thereby compute
the transfer function of the software-defined policy.  Previous work
has shown that a wide range of micropolicies can be defined and
enforced efficiently \rb{cite}.
%
\rb{Maybe go into more detail, examples later\ldots or try to segue into stack
policies}

\paragraph*{A conservative policy}
%
The first micropolicy we present to enforce the definition of stack
safety developed in \cref{sec:lse-and-wbcf} is \leo{a small variation
  of?}  the Depth Isolation policy presented by
\citet{DBLP:conf/sp/RoesslerD18}. Their policy tags the stack memory
with ownership information associated to each stack frame (and to each
separate object inside that frames, a more finely grained access
control that we do not need to consider here), and registers holding
stack pointers with matching access permission information. Beyond
those capabilities, the compiler generates designated sequences of
instructions with special tags that grant them permission to manage
the stack, say during calls and returns. \rb{Conceivably on alloc
  operations, etc., which we have not yet mentioned.}

Our micropolicy works as follows. We assume for ease of exposition that no
arguments are passed on the stack \rb{Come back to this later?}. Initially, the
entire stack is tagged as unused with an $\tagNoDepth$ tag \rb{actually, the
``initial frame'' should be tagged with its proper depth}, and the {\PCname}
carries in its tag at all times the depth of the current activation, starting
with a $\tagPCDepth{0}$ tag. The current activation depth needs to change
exactly at those points in the program where the contour of the stack safety
property changes:

@@

\item From caller to callee, when the machine executes an instruction tagged
  with $\tagCall$ as belonging to the call map. At this point, the current
  {\PCname} depth $n$ is incremented to $\tagPCDepth{n + 1}$; on function entry,
  the locations of the new frame is marked $\tagStackDepth{n + 1}$.

@@

This conservative policy does not only satisfy the stack safety above,
but also a provably (in Coq) stronger one that enforces local state
encapsulation throughout every calls execution. To encode that
property, we not only keep track of the execution of one machine
(whose trace we're trying to characterize as safe), but of an entire
stack of machine variations corresponding to nested calls of the
original machine. For each step of the original machine, we take a
step in every variant machine (thus, keeping them in lockstep
execution) and enforce a single-step variant of integrity and
confidentiality tests. In addition, every time a call is made we push
a variation of the original machine state at the time of that call on
top of the variant stack (just like the stack confidentiality property
enforced trace confidentiality for every transition), and every time
the machine returns from a call we pop the top variation (to account
for stack confidentiality taking the prefix of the transition trace
until its return).

@@

%%   \item
%% %
%%     If the machine takes a \emph{call step} and $\contour$ is the contour at
%%     the call then,
%% %
%%     (1) the variant stack test $\testVS{(\mach, \pol)}{\vs}$ must be
%%     satisfied,
%% %
%%     (2) the stack integrity test $\testInt{\contour}{\mach}{\mach'}$ is
%%     satisfied and the stack confidentiality test $\testConf{\mach}{\vm{\contour}}$
%%     is satisfied with resepct to an arbitrary $\contour$-variant
%%     of $\mach$,
%% =======
We also need two auxilliary notations that relate to variant stacks.
% 
First, strong safety will check that the basic tests above hold for every
machine in the variant stack.  Given a machine that steps $(\mach,
\pol) \hookrightarrow_P (\mach', \pol', \obs)$, say that a variant
stack $\vs$ safisfies the \emph{variant stack test} if for each
variant element $\vm{C}$ (which is a variation of $\mach$ for $C$),
$\vm{C}$ can take a step, denoted $\vm{C} \stepstoobs{\obs} \mach'$,
while satisfying the stack integrity test $\testInt{C}{\mach}{\mach'}$
and stack confidentiality test $\testConf{\mach}{\mach'}$ with regard
to the original machine state. We will write $\testVS{(\mach,
  \pol)}{\vs}$ for this compound check.
%% >>>>>>> 1886f7a7bf858ce0c9518e407d9481518a4f028d
%
Second, we will write $\vs \stepsto \vs'$ if every variant machine
in $\vs$ steps to the corresponding machine in $\vs'$.
%

@@

\item
  If the machine takes a \emph{call step} and $\contour$ is the contour at
  the call
  \[(\mach, \pol) \hookrightarrow_P (\mach', \pol', \obs) \wedge \callmap(\mach) = n,\]
%
  then the variant stack test $\testVS{(\mach, \pol)}{\vs}$ must be
  satisfied:
  \[\testVS{(\mach, \pol)}{\vs},\]
%
%    (2) the stack integrity test $\testInt{\contour}{\mach}{\mach'}$ is
%    satisfied and the stack confidentiality test $\testConf{\mach}{\vm{\contour}}$
%    is satisfied with regard to an arbitrary $\contour$-variant
%    of $\mach$,
%
  and trace strong stack safety must be coinductively satisfied
  for the remainder of the trace and for every variant stack
  that is the result of adding a variation of $\mach'$ with
  respect to $C$ at the top of the stack $\vs'$ that $\vs$ steps
  to:
  \[\testProp{}{((\mach', \pol', \obs)\MPT)}{(\vm{\contour} : \vs')}\]

\item
%
  If the machine $\mach$ takes a \emph{return step},
  \[(\mach, \pol) \hookrightarrow_P (\mach', \pol', \obs) \wedge \retmap(\mach),\]
  %
  the top of the variant stack also returns,
  the tail of $\vs$ satisfies the stack safety test
  \[\testVS{(\mach, \pol)}{\mathit{tail}(\vs)},\]
  %
  and trace strong stack safety must be coinductively satisfied for
  the remainder of the trace and for the tail of the varant stack
  that $\vs$ steps to:
  \[\testProp{}{((\mach', \pol', \obs)\MPT)}{(\mathit{tail}(\vs'))}\]

>>>>

The intuition behind \citeauthor{Skorstengaard+19}'s use of the term {\em
  local state encapsulation} is that callers should be protected from
callees: a callee should not be able to access or overwrite private data
in the caller's stack frame. We formalize this intuition as a security
property with both confidentiality and integrity components.
%
We reflect this intuition in a formal
{\em stack confidentiality} property, which promises that a callee's
observable behavior is not affected by memory outside its stack frame
(e.g., a caller's private data or uninitialized memory).  Conversely, {\em stack
  integrity} promises that, if a function call ever returns, the caller's
stack data at the point of return will be identical to what it was before
the call.

@@

To phrase these two aspects of stack safety in terms of security
properties \leo{in terms of security properties or ``using security
  language''?}  requires an explicit attacker model: a clear
understanding of who is being protected and from what. Most existing
stack protection mechanisms deal with scenarios where a callee gets
influenced by external inputs and ends up executing potentially
malicious code. To avoid the details of exactly how this happens and
keep the presentation clear, we pick an even stronger model: when
stating the protections that a function can rely on when it makes a
call, we assume that the callee can execute arbitrary instructions
until it returns.

@@

Enforcement mechanisms are modeled as {\em policies} that add some auxiliary
policy state to the machine state and a partial policy step function that
refines the machine step function by allowing or disallowing a step to be taken
based on its auxiliary state. This model will be made more precise in
\cref{sec:prelim}. It is flexible enough to capture any kind of mechanism used
for dynamic enforcement of properties of interest, including the following
representative families of policies.

@@

\paragraph*{Null policy}
%
The simplest possible policy is the neutral policy that carries no information
and does nothing. Its set of policy states is the unit type, and its policy step
function is the constant unit function, thereby allowing all calls to the machine
step function. Its initialization function is defined on all states. This
corresponds to the lifting of our machine model to the policy setting. Machines
that fulfill a property by construction will also fulfill that property with the
null policy.

\paragraph*{Software-only policies}
%
Some policies are implemented in software, by modifying code in the compiler to
(e.g.) perform bounds checking \citep{NagarakatteZMZ09} or to insert stack
canaries \citep{Cowan+98}. These approaches are represented as policies that
accept only initial machine states that are well-formed for the policy (due to
being produced by the compiler.) The set of policy states is again the unit type
and the step function constant. Property enforcement comes from the fact that
only ill-formed states can induce a trace that violates the property.

@@

\paragraph*{Capability machines}
%
Other policies are implemented in hardware or built on such hardware mechanisms.
Capability machines like CHERI \citep{Woodruff+14} extend conventional
architectures to support efficient and fine-grained control over memory
accesses, which can be used to implement security policies involving, for
example, compartmentalization or the stack protection policies of
\citet{Skorstengaard+19,Skorstengaard+19b}. Mapping the underlying protection
mechanisms to policy states and step function would be a delicate operation, as
those are tightly coupled to the hardware design. However, separating the
enforcement aspects of a concrete policy (such as StkTokens, where policy state
would encompass details like the abstract call stack, the capability sealing
scheme, etc. \rb{is this digging too deep? or on the contrary, say more?}) from
the machine would be relatively straightforward.

@@

In this section we describe our machine model. Our definition of stack
safety is generic and this section discusses an abstract interface to
our machine model and some kind of static or dynamic, software- or
hardware-enforced security monitor. We will refer back to the running
example of \cref{sec:running-example} for concreteness, and then,
in~\cref{sec:enforcement}, extend this model with a dynamic policy
enforcement mechanism based on
\citet{DBLP:conf/sp/RoesslerD18}.\apt{needs a relook}

@@

In our model, the machine step function is total---a program does not halt
on its own, but may instead enter a silent terminal loop. Runs are only cut
short by the policy step function. This means that software-only policies that
can catch errors must be modeled as reaching an error state\apt{??}, not 
halting outright.\apt{Is this our current thinking?}
\rb{And if so, should the running example loop at the end, or do we assume
that is exactly what happens after main is done?}

@@

In this section we formalize our definition of stack safety. We
begin with our model of \leo{compiler-provided?} knowledge about the
program structure, and then dive into the formal definitions of local state
encapsulation and well-bracketed control flow that together constitute
stack safety.

@@

\definition%{Stack Confidentiality}
%Let $\callmap$ be a call map and $\codemap$ a function map.
We will say that a system satisfies {\em stack confidentiality} with respect to
\(\callmap\) in a similar way to stack integrity. For every initial state
\(\mach_0\), we take the induced trace \(\mpstate{\mach_0}{\pinit(\mach_0)}
\hookrightarrow \MPT\). Then for any \(n\)-transition
\(\mathit{Tr}_n ~ \MPT ~ \mpstate{\mach_c}{\pol_c} ~
\mpstate{\mach_e}{\pol_e}\), we take the prefix trace until return:
%
\[\mpstate{\mach_e}{\pol_e} \hookrightarrow \MPT_{\mathit{call}} | \ret{\mach_c}\]

We also take the corresponding trace from any variant \(\mach_e' 
\approx_{\mathit{Cof}(\mach_c,n)} \mach_e\):
%
\[\mach_e' \hookrightarrow \machT_{\mathit{call}} | \ret{\mach_c}\]

@@

\rb{Wondering if we should use sub-sections here, too.}

@@

\paragraph*{Micropolicies}
%
The enforcement policy of \citeauthor{DBLP:conf/sp/RoesslerD18} relies
on a programmable, tag-based reference monitor that runs alongside the
program.  To control this monitor, they use a programming model that
allows fine-grained manipulation of metadata tags to encode so-called
\emph{micropolicies}~\citep{pump_oakland15}.

In such a system, all memory addresses and registers (including the program
counter) are enriched with an abstract metadata tag, which can represent an
arbitrary data structure. A micropolicy is defined as a set of tags, and a
collection of software-defined rules that represent the transfer function of the
policy. Each rule inspects the tags associated to the inputs of an instruction
(PC, registers, memory) as well as the instruction opcode itself, and produces
one of two outcomes: either the instruction is allowed to execute (and generates
and assigns tags for the new PC and the result of the operation), or the machine
fail-stops with a policy violation. During program execution, micropolicy rules
are checked before each machine step.

Efficient execution of these micropolicies relies on hardware implementations,
such as the PUMP architecture~\citep{pump_oakland15}. This hardware incorporates
a monitor rule cache to allow quick retrieval of transfer function outputs for
mapped inputs. Cache misses are detected and trapped to software handlers that
use the software rules of the policy to compute the relevant parts of the
transfer function. Previous work has shown that a wide range of micropolicies
can be defined and enforced efficiently in these machines
\citep{pump_oakland2015}.
%
\rb{Maybe go into more detail, examples later\ldots or try to segue into stack
policies}

\paragraph*{A conservative policy}
%
The first micropolicy we present to enforce the definition of stack
safety developed in \cref{sec:lse-and-wbcf} is \leo{a small variation
  of?}  the Depth Isolation policy presented by
\citet{DBLP:conf/sp/RoesslerD18}. Their policy tags the stack memory
with ownership information associated to each stack frame (and to each
separate object inside that frame, a more finely grained access
control that we do not need to consider here), and registers holding
stack pointers with matching access permission information. Beyond
those capabilities, the compiler generates designated sequences of
instructions with special tags that grant them permission to manage
the stack, say during calls and returns. \rb{Conceivably on alloc
  operations, etc., which we have not yet mentioned.}

Our micropolicy works as follows. We assume for ease of exposition that no
arguments are passed on the stack \rb{Come back to this later?}. Initially, the
entire stack is tagged as unused with an $\tagNoDepth$ tag \rb{actually, the
``initial frame'' should be tagged with its proper depth}, and the {\PCname}
carries in its tag the depth of the current activation at all times, starting
with a $\tagPCDepth{0}$ tag. The current activation depth needs to change
exactly at those points in the program where the contour of the stack safety
property changes:

@@

\item From caller to callee, when the machine executes an instruction tagged
  with $\tagCall$ as belonging to the call map. At this point, the current
  {\PCname} depth $n$ is incremented to $\tagPCDepth{n + 1}$; on function entry,
  the locations of the newly allocated frame are marked $\tagStackDepth{n + 1}$.

@@

This conservative policy satisfies not only the stack safety above,
but also a provably (in Coq) stronger one that enforces local state
encapsulation throughout every call's execution. To encode that
property, we not only keep track of the execution of one machine
(whose trace we are trying to characterize as safe), but of an entire
stack of machine variations corresponding to nested calls of the
original machine. For each step of the original machine, we take a
step in every variant machine (thus, keeping them in lockstep
execution) and enforce a single-step variant of integrity and
confidentiality tests. In addition, every time a call is made we push
a variation of the original machine state at the time of that call on
top of the variant stack (just like the stack confidentiality property
enforced trace confidentiality for every transition), and every time
the machine returns from a call we pop the top variation (to account
for stack confidentiality taking the prefix of the transition trace
until its return).

@@

%%   \item
%% %
%%     If the machine takes a \emph{call step} and $\contour$ is the contour at
%%     the call then,
%% %
%%     (1) the variant stack test $\testVS{(\mach, \pol)}{\vs}$ must be
%%     satisfied,
%% %
%%     (2) the stack integrity test $\testInt{\contour}{\mach}{\mach'}$ is
%%     satisfied and the stack confidentiality test $\testConf{\mach}{\vm{\contour}}$
%%     is satisfied with resepct to an arbitrary $\contour$-variant
%%     of $\mach$,
%% =======
We also need two auxilliary notations that relate to variant stacks.
%
First, strong safety will check that the basic tests above hold for every
machine in the variant stack.  Given a machine that steps $(\mach,
\pol) \stepstopol{\obs} (\mach', \pol')$, say that a variant
stack $\vs$ safisfies the \emph{variant stack test} if for each
variant element $\vm{C}$ (which is a variation of $\mach$ for $C$),
$\vm{C}$ can take a step, denoted $\vm{C} \stepstoobs{\obs} \mach'$,
while satisfying the stack integrity test $\testInt{C}{\mach}{\mach'}$
and stack confidentiality test $\testConf{\mach}{\mach'}$ with regard
to the original machine state. We will write $\testVS{(\mach,
  \pol)}{\vs}$ for this compound check.
%% >>>>>>> 1886f7a7bf858ce0c9518e407d9481518a4f028d
%
Second, we will write $\vs \stepsto \vs'$ if every variant machine
in $\vs$ steps to the corresponding machine in $\vs'$.
%

@@

\item
  If the machine takes a \emph{call step} and $\contour$ is the contour at
  the call \rb{Corrected the step function here, or should it be
  $\hookrightarrow$ and give the rest of the induced trace?}
  \[(\mach, \pol) \stepstopol{\obs} (\mach', \pol') \wedge \callmap(\mach) = n,\]
%
  then the variant stack test $\testVS{(\mach, \pol)}{\vs}$ must be
  satisfied:
  \[\testVS{(\mach, \pol)}{\vs},\]
%
%    (2) the stack integrity test $\testInt{\contour}{\mach}{\mach'}$ is
%    satisfied and the stack confidentiality test $\testConf{\mach}{\vm{\contour}}$
%    is satisfied with regard to an arbitrary $\contour$-variant
%    of $\mach$,
%
  and trace strong stack safety must be coinductively satisfied
  for the remainder of the trace and for every variant stack
  that is the result of adding a variation of $\mach'$ with
  respect to $C$ at the top of the stack $\vs'$ that $\vs$ steps
  to:
  \[\testProp{}{((\mach', \pol', \obs)\MPT)}{(\vm{\contour} : \vs')}\]

\item
%
  If the machine $\mach$ takes a \emph{return step},
  \[(\mach, \pol) \stepstopol{\obs} (\mach', \pol') \wedge \retmap(\mach),\]
  %
  the top of the variant stack also returns,
  the tail of $\vs$ satisfies the stack safety test
  \[\testVS{(\mach, \pol)}{\mathit{tail}(\vs)},\]
  %
  and trace strong stack safety must be coinductively satisfied for
  the remainder of the trace and for the tail of the varant stack
  that $\vs$ steps to:
  \[\testProp{}{((\mach', \pol', \obs)\MPT)}{(\mathit{tail}(\vs'))}\]
